Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-09-05 writeup/conStruct.tex
G's comments

\begin{abstract}
One of the classic problems in population genetics is the characterization 
of discrete population structure when the genotyped samples also show 
continuous patterns of genetic differentiation.
Especially when sampling is discontinuous, 
clustering or assignment methods may incorrectly ascribe differentiation 
due to continuous processes (e.g., isolation by distance) 
to discrete processes, such as geographic, ecological, or reproductive barriers 
between populations.
This is partly a result of the difficulty of sampling uniformly and continuously 
across the range of a population or species, 
but more, it reflects a shortcoming of current methods for inferring and 
visualizing population structure from genetic data in the face of data 
that are characterized by both continuous and discrete population structure.
Here, we present a novel statistical framework for the simultaneous inference 
of continuous and discrete patterns of population structure.
The method estimates ancestry proportions for each 
sample from a set of discrete population layers, 
and, within each layer, estimates a rate at which relatedness decays with distance.
This model explicitly addresses the ``clines vs. clusters" problem in 
modeling population genetic variation by jointly accommodating both 
continuous and discrete patterns of differentiation. 
We demonstrate the utility of this approach using both simulations and empirical applications.
\end{abstract}

\newpage
\section*{Introduction}
A fundamental quandary in the description of biological diversity is
the fact that diversity shows both discrete and continuous patterns. 
For example, there is disagreement on species concepts because the
process of speciation is usually gradual, and so there is no set point
in the continuous divergence of populations when they unambiguously become discrete species.
This issue extends below the species level, as patterns of phenotypic
and genetic diversity within and among populations are shaped
by continuous migration and drift, as well as by more discrete events, 
such as geographic barriers, rapid expansions, bottlenecks, and rare
long distance migration.

From a practical standpoint, even while acknowledging continuous
processes, we often need to identify somewhat separable populations 
from which individuals are sampled \citep{wright1949genetical}.
Delineating populations is useful for systematics and for
informing conservation priorities \citep{Moritz1994,Waples_1998,Moritz_etal_2002}.
Furthermore, we often need to identify sets of indivduals resulting from reasonably coherent
evolutionary histories for downstream analyses to learn about population history and adaptation.

There have been many methods proposed to characterize population
genetic structure,
including generating population phylogenies \citep{CavalliSforza1975, treemix},
$k$-means clustering analyses performed on dimensionality-reduction approaches \citep{meirmans2009genodive}, 
such as principal components analysis \citep{menozzi1978synthetic,novembre_interpreting_2008, price2006eigenstrat},
and model-based clustering approaches \citep[e.g.][]{STRUCTURE, falush2003, hubisz2009,ADMIXTURE, FINESTRUCTURE, fastStructure, huelsenbeck2007inference, Corander2003,TESS,geneland}, among others.
Here, we will focus on model-based clustering, the most widely used class of approaches for population delineation.
Like the method we present here, these are based on a statistical model,
but do not perform inference directly with a demographic model.
Each of these methods perform best in particular situations,
but all can give misleading results when applied to data that show a continuous pattern of isolation by distance 
\citep{Wright1943, novembre_interpreting_2008, Frantz2009}.

A pattern of isolation by distance (IBD) arises when migration in a population is spatially limited \citep{Wright1943},
and has been modeled in both spatially continuous frameworks \citep{malecot1969mathematics,Slatkin1985}, 
in which an individual's parents are found within some radius of that individual's location,
and discrete, ``stepping-stone'' frameworks \citep{kimura1964stepping},
in which migration occurs only between neighboring demes.
Given the generality of the circumstances that generate a pattern of isolation by distance, 
it is unsurprising that IBD is very widespread in nature \citep{meirmans2012,Sexton_etal_2014}.

The ubiquity of isolation by distance presents a challenge for models of discrete population structure,
as it is frequently difficult to determine whether observed patterns of genetic variation are 
continuously distributed across a landscape, or instead are partitioned in discrete clusters.
This problem can be compounded if sampling is done unevenly or discretely across a population or species' range,
and has given rise to a debate in the population genetic literature about whether 
individuals are best described as being drawn from continuous clines or discrete clusters 
\citep[e.g.][]{SerrePaabo2004,rosenberg2005clines}.
%As a further complication, if local genetic drift is strong,
%continous models are expected to give rise to spatially patchy patterns
%\citep{SLFV,bramson1989crabgrass}.

@@

\paragraph{Data}
The statistical framework of our approach is conceptually similar to that in \cite{Wasser2004}, \cite{BEDASSLE}, and \cite{spacemix},
although we use a somewhat different summary statistic than in this previous work.
The model works with allele frequencies at $L$ unlinked, bi-allelic single nucleotide polymorphisms (SNPs) genotyped across $N$ samples.
Each ``sample'' may be a collection of diploids, or frequencies estimated from pooled sequencing.
We write $N_i$ for the number of chromosomes sequenced in the $i^\text{th}$ sample.
The \emph{sample frequency} at locus $\ell$ in sample $n$, denoted $f_{n,\ell}$, 
is calculated by first arbitrarily choosing one of the observed alleles at locus $\ell$ to count, 
then dividing the number of observations of that counted allele by the total number of chromosomes genotyped at that locus
in sample $n$.
The choice of allele does not affect subsequent calculations, and so may be arbitrary.
We then calculate the \emph{allelic covariance} between samples $i$ and $j$, denoted $\widehat{\Omega}_{i,j}$,
as the expected covariance of distinct individual alleles chosen from each of the two samples at a random locus,
choosing a random allele to code as `1' (and the other as `0').
Concretely, this is
\begin{equation}
\widehat{\Omega}_{i,j} = 
    \frac{1}{L} \sum_{\ell=1}^L \left( (f_{i,\ell}-1/2) (f_{j,\ell}-1/2) - \delta_{i,j} f_{i,\ell}^2 / (N_i-1) \right),
\label{allelic_covariance}
\end{equation}
where $\delta_{i,j}=1$ if $i=j$ and is 0 otherwise.
Although we describe this as a covariance between individually drawn alleles,
$\widehat{\Omega}_{i,j}$ is in fact also the covariance between the allele frequencies
of a randomly chosen allele in samples $i$ and $j$, as long as $i \neq j$.
The diagonal (where $i=j$) does not have this interpretation.

In addition to these differences on the diagonal,
this differs from the usual ``genetic covariance'' \citep{mcvean_genealogical_2009}
in that (a) we do not subtract locus means,
and so to make the statistic invariant under choices of reference allele, 
(b) we randomly choose an allele to count at each locus.
We discuss the derivation of Eqn. \eqref{allelic_covariance} further in the Appendix (\secref{allelic_cov}).
As noted in \cite{EEMS},
this covariance has a close relationship to the pairwise genetic distance:
if $D_{i,j}$ is the mean density of sites at which random samples from $i$ and $j$ differ at a randomly chosen locus,
then $\Omega_{i,j} = (1 - 2 D_{i,j})/4$.
Therefore, unlike genetic covariance,
this allelic covariance is affected in shape by singleton sites,
so it may be advisable to filter these prior to analysis
if these are likely to contain a large percentage of errors.

\paragraph{Continuous and discrete differentiation}
Clustering approaches to describing genetic variation
are useful because population history can often be meaningfully described on a coarse scale
by interactions between discrete ``populations''
whose relationships are delimited by patterns of glaciation,
large-scale migration,
mountain ranges, and the like.
Here we add a spatial component within each such discrete historical component,
which we refer to as a set of ``layers'' that overlay the modern map.
We imagine each layer as a geographically distributed population,
and that each sample is composed of a mixture of contributions from each of these layers.
These layers thus take the place of ``clusters'' or ``populations'' in clustering methods,
but we do not adopt these terms as they connote a lack of geographical separation within each.

@@

To allow flexibility in the form of the decay of allelic covariance with geographic distance within each cluster, 
we define the covariance within cluster $k$ between samples $i$ and $j$ to be:
\begin{equation}
G^{(k)}_{i,j} 
    = 
    \alpha^{(k)}_0 \left( \text{exp} \left( -(\alpha^{(k)}_D D_{i,j}) ^ {\alpha^{(k)}_2}	\right) \right) + \phi^{(k)}
\label{within_cluster_covariance}
\end{equation}
where the superscript $(k)$ denotes parameters specific to the $k$th cluster.
The quantity $D_{i,j}$ is the observed geographic distance between samples $i$ and $j$,
the $\alpha^{(k)}$ parameters control the shape of the decay of covariance with distance in the cluster,
and $\phi^{(k)}$ is a parameter that describes the background covariance within the cluster. 
If two samples draw 100\% of their ancestry from cluster $k$, then their covariance under the model is $G^{(k)}_{i,j}$;
if they are furthermore geographically very close ($D_{i,j}=0$)
they will have covariance $\alpha^{(k)}_0 +  \phi^{(k)}$.
If the geographic distance between them is very large, 
their covariance will be equal to to the background level $\phi^{(k)}$ within the cluster.
The ``shared drift'' parameter $\phi^{(k)}$ can be interpreted as the branch length 
connecting the $k$th population to the population ancestral to all modeled
layers \citep{patterson_ancient_2012, peter_fstats}.

We then allow samples to draw their ancestry from more than one cluster.
The ``admixture'' proportion of the $i$th sample in the $k$th cluster, denoted $w^{(k)}_i$,
gives the proportion of alleles from sample $i$ that derive from
cluster $k$ (and so $\sum_K w^{(k)}_i =1$).
A visual representation of the method mechanics is shown in Fig \ref{schematic}.

>>>>

\begin{abstract}
\gc{A classic problem in population genetics is the characterization 
of discrete population structure in the presence of} 
continuous patterns of genetic differentiation.
Especially when sampling is discontinuous, 
clustering or assignment methods may incorrectly ascribe differentiation 
due to continuous processes (e.g., isolation by distance) 
to discrete processes, such as geographic, ecological, or reproductive barriers 
between populations.
This is partly a result of the difficulty of sampling uniformly and continuously 
across the range of a population or species, 
but more, it reflects a shortcoming of current methods for inferring and 
visualizing population structure from genetic data in the face of data 
that are characterized by both continuous and discrete population structure.
Here, we present a novel statistical framework for the simultaneous inference 
of continuous and discrete patterns of population structure.
The method estimates ancestry proportions for each 
sample from a set of discrete population layers, 
and, within each layer, estimates a rate at which relatedness decays with distance.
This model explicitly addresses the ``clines vs. clusters" problem in 
modeling population genetic variation by jointly accommodating both 
continuous and discrete patterns of differentiation. 
We demonstrate the utility of this approach using both simulations and empirical applications.
\end{abstract}

\newpage
\section*{Introduction}
A fundamental quandary in the description of biological diversity is
the fact that diversity shows both discrete and continuous patterns. 
For example, \gc{reasonable people can disagree whether two
  populations are separate species, because the} %there is disagreement on species concepts because the
process of speciation is usually gradual, and so there is no set point
in the continuous divergence of populations when they unambiguously become discrete species.
\gc{The issue of identifying meaningful biological subunits} extends below the species level, as patterns of phenotypic
and genetic diversity within and among populations are shaped
by continuous migration and drift, as well as by more discrete events, 
such as geographic barriers, rapid expansions, bottlenecks, and rare
long distance migration.

From a practical standpoint, even while acknowledging continuous
processes, we often need to identify somewhat separable populations 
from which individuals are sampled \citep{wright1949genetical}.
Delineating populations is useful for systematics and for
informing conservation priorities \citep{Moritz1994,Waples_1998,Moritz_etal_2002}.
Furthermore, we often need to identify \gc{subsets} of indivduals resulting from reasonably coherent
evolutionary histories for downstream analyses to learn about population history and adaptation.

There have been many methods proposed to characterize population
genetic structure,
including generating population phylogenies \citep{CavalliSforza1975, treemix},
\gc{dimensionality-reduction} approaches \citep{meirmans2009genodive},   %$k$-means clustering analyses performed
such as principal components analysis \citep{menozzi1978synthetic,novembre_interpreting_2008, price2006eigenstrat},
and model-based clustering approaches \citep[e.g.][]{STRUCTURE, falush2003, hubisz2009,ADMIXTURE, FINESTRUCTURE, fastStructure, huelsenbeck2007inference, Corander2003,TESS,geneland}, among others.
Here, we will focus on model-based clustering, the most widely used class of approaches for population delineation.
\gc{We need more on CLUSTERING approaches here. SOMETHING LIKE:
These clustering methods model each individual's genotypes as random
draws from a set of underlying, unobserved
population clusters, each with a characteristic set of allele
frequencies. 
These underlying frequencies are identical for all individuals
assigned to a cluster, regardless of their spatial location. 
Spatial information has been incorporated in some of these methods, for
example in placing priors on cluster memberships \citep{geneland}, but
this does not address the underlying issue that allele frequencies are
assumed constant in a cluster across the species range.  
}
Like the method we present here, these are based on a statistical model,
but do not perform inference directly with a demographic model.
Each of these methods perform best in particular situations,
but all can give misleading results when applied to data that show a continuous pattern of isolation by distance 
\citep{Wright1943, novembre_interpreting_2008, Frantz2009}.

A pattern of isolation by distance (IBD) arises when migration in a
population is spatially limited, \gc{and slow enough betweeen distant
locations that genetic drift can build up differentiation}
\citep{Wright1943}. IBD has been modeled in both spatially continuous frameworks \citep{malecot1969mathematics,Slatkin1985}, 
in which an individual's parents are found within some radius of that individual's location,
and discrete, ``stepping-stone'' frameworks \citep{kimura1964stepping},
in which migration occurs only between neighboring demes.
Given the generality of the circumstances that generate a pattern of isolation by distance, 
it is unsurprising that IBD is very widespread in nature \citep{meirmans2012,Sexton_etal_2014}.

The ubiquity of isolation by distance presents a challenge for models of discrete population structure,
as it is frequently difficult to determine whether observed patterns of genetic variation are 
continuously distributed across a landscape, or instead are partitioned in discrete clusters.
This problem can be compounded if sampling is done unevenly or discretely across a population or species' range,
and has given rise to a debate in the population genetic literature
about \gc{how best to describe sets of individuals using continuous clines and} discrete clusters 
\citep[e.g.][]{SerrePaabo2004,rosenberg2005clines}.
%As a further complication, if local genetic drift is strong,
%continous models are expected to give rise to spatially patchy patterns
%\citep{SLFV,bramson1989crabgrass}.

@@

\paragraph{Data}
The statistical framework of our approach is conceptually similar to that in \cite{Wasser2004}, \cite{BEDASSLE}, and \cite{spacemix},
although we use a somewhat different summary statistic than in this previous work.
The model works with allele frequencies at $L$ unlinked, bi-allelic single nucleotide polymorphisms (SNPs) genotyped across $N$ samples.
Each ``sample'' may \gc{be a single individual, a collection of
  individuals from a location}, or frequencies estimated from pooled sequencing.
We write $N_i$ for the number of chromosomes sequenced in the $i^\text{th}$ sample.
The \emph{sample frequency} at locus $\ell$ in sample $n$, denoted $f_{n,\ell}$, 
is calculated by first arbitrarily choosing one of the observed alleles at locus $\ell$ to count, 
then dividing the number of observations of that counted allele by the total number of chromosomes genotyped at that locus
in sample $n$.
The choice of allele does not affect subsequent calculations, and so may be arbitrary.
We then calculate the \emph{allelic covariance} between samples $i$ and $j$, denoted $\widehat{\Omega}_{i,j}$,
as the expected covariance of distinct individual alleles chosen from each of the two samples at a random locus,
choosing a random allele to code as `1' (and the other as `0').
Concretely, this is
\begin{equation}
\widehat{\Omega}_{i,j} = 
    \frac{1}{L} \sum_{\ell=1}^L \left( (f_{i,\ell}-1/2) (f_{j,\ell}-1/2) - \delta_{i,j} f_{i,\ell}^2 / (N_i-1) \right),
\label{allelic_covariance}
\end{equation}
where $\delta_{i,j}=1$ if $i=j$ and is 0 otherwise.
Although we describe this as a covariance between individually drawn alleles,
$\widehat{\Omega}_{i,j}$ is in fact also the covariance between the allele frequencies
of a randomly chosen allele in samples $i$ and $j$, as long as $i \neq j$.
The diagonal (where $i=j$) does not have this interpretation, \gc{and
  reflects the covariance between indivdual alleles drawn from within
  the population}.

\gc{This definition of convariance} differs from the usual ``genetic covariance'' \citep{mcvean_genealogical_2009}
in that (a) we do not subtract locus means,
and so to make the statistic invariant under choices of reference allele, 
(b) we randomly choose an allele to count at each locus.
We discuss the derivation of Eqn. \eqref{allelic_covariance} further in the Appendix (\secref{allelic_cov}).
As noted in \cite{EEMS},
this covariance has a close relationship to the pairwise genetic distance:
if $D_{i,j}$ is the mean density of sites at which random samples from $i$ and $j$ differ at a randomly chosen locus,
then $\Omega_{i,j} = (1 - 2 D_{i,j})/4$.
Therefore, unlike genetic covariance,
this allelic covariance is \gc{more} affected in shape by singleton sites,
so it may be advisable to filter these prior to analysis
if these are likely to contain a large percentage of errors.

\paragraph{Continuous and discrete differentiation}
Clustering approaches to describing genetic variation
are useful because population history can often be meaningfully described on a coarse scale
by interactions between discrete ``populations''
whose relationships are delimited by patterns of glaciation,
large-scale migration,
mountain ranges, and the like.
Here we add a spatial component within each such discrete historical component,
which we refer to as a set of ``layers'' that overlay the modern map.
We imagine each layer as a geographically distributed population,
\gc{each layer extends over the entire sampled range of the
  populations}. 
Each sample is composed of a mixture of contributions from each of
these layers. \gc{See Figure \ref{schematic} for a cartoon representation
of our model.} 
These layers thus take the place of ``clusters'' or ``populations'' in clustering methods,
but we do not adopt these terms as \gc{terms like ``spatial'' cluster
  suggest a clustering in space while our layers (and samples from
  them) may be present at all sampled locations.}

%they connote a lack of geographical separation within each.

@@

To allow flexibility in the form of the decay of allelic covariance with geographic distance within each cluster, 
we define the covariance within cluster $k$ between samples $i$ and $j$ to be:
\begin{equation}
G^{(k)}_{i,j} 
    = 
    \alpha^{(k)}_0 \left( \text{exp} \left( -(\alpha^{(k)}_D D_{i,j}) ^ {\alpha^{(k)}_2}	\right) \right) + \phi^{(k)}
\label{within_cluster_covariance}
\end{equation}
where the superscript $(k)$ denotes parameters specific to the $k$th cluster.
The quantity $D_{i,j}$ is the observed geographic distance between samples $i$ and $j$,
the $\alpha^{(k)}$ parameters control the shape of the decay of
covariance with distance in the cluster.
\gc{Our choice of a powered-exponential decay, as parameterized by the
  $\alpha$s, is a flexible, and standard choice in spatial statistics
  REF, and is not chosen to match  a particular population genetics model. 
The} $\phi^{(k)}$ is a parameter that describes the background covariance within the cluster. 
If two samples draw 100\% of their ancestry from cluster $k$, then their covariance under the model is $G^{(k)}_{i,j}$;
if they are furthermore geographically very close ($D_{i,j}=0$)
they will have covariance $\alpha^{(k)}_0 +  \phi^{(k)}$.
If the geographic distance between them is very large, 
their covariance will be equal to to the background level $\phi^{(k)}$ within the cluster.
The ``shared drift'' parameter $\phi^{(k)}$ can be interpreted as the branch length 
connecting the $k$th population to the population ancestral to all modeled
layers \citep[see for example ][]{patterson_ancient_2012, peter_fstats}.

We then allow samples to draw their ancestry from more than one cluster.
The ``admixture'' proportion of the $i$th sample in the $k$th cluster, denoted $w^{(k)}_i$,
gives the \gc{genome-wide} proportion of alleles from sample $i$ that derive from
cluster $k$ (and so $\sum_K w^{(k)}_i =1$).
A visual representation of the method mechanics is shown in Fig \ref{schematic}.
